A Simplified Plackett-Luce Likelihood Function for Rank Orders

### 2025-10-02 inst/exponential_BT.Rmd
more work

```{r makeII, cache=TRUE}
r_try <- 1:9
x_try <- seq(from = 0.1, to = 1, by = 0.1)
jj <- as.matrix(expand.grid(r_try, x_try))
II <- t(apply(jj, 1, function(v){getprob(9, v[1], v[2], N=1e4)}))
head(II)
```

```{r makesupp, cache=TRUE}
r_true <- 3
x_true <- 0.85
o <- getprob(9, r_true, x_true, 17)  # observations
o[1:2]
o[-(1:2)]

support_func <- function(o, v){
    v <- v[-(1:2)]
    v <- v/sum(v)
    dmultinom(o[-(1:2)], prob=v, log=TRUE)
}
support <- rep(0,nrow(II))
for(i in seq_len(nrow(II))){
    support[i] <- support_func(o,II[i,])
}
```

```{r}
support <- support-max(support)
support <- pmax(support, -10)
support  <- matrix(support,length(r_try),length(x_try))
contour(r_try, x_try, support, xlab="r", ylab="x", levels = -(1:5)*2)
abline(v = r_true)
abline(h = x_true)
abline(v = 1:9, col = 'gray', lty = 3, lwd = 0.3)
filled.contour(r_try, x_try, support, xlab = "r", ylab = "x")
```

>>>>

```{r makeII, cache=TRUE}
r_try <- 1:9
x_try <- seq(from = 0.1, to = 1, by = 0.05)
jj <- as.matrix(expand.grid(r_try, x_try))
II <- t(apply(jj, 1, function(v){getprob(9, v[1], v[2], N=1e4)}))
head(II)
```

```{r makesupp, cache=TRUE}
r_true <- 3
x_true <- 0.85
set.seed(0)
o <- getprob(9, r_true, x_true, 100)  # observations
o[1:2]
o[-(1:2)]

support_func <- function(o, v){
    v <- v[-(1:2)]
    v <- v/sum(v)
    dmultinom(o[-(1:2)], prob=v, log=TRUE)
}
support <- rep(0,nrow(II))
for(i in seq_len(nrow(II))){
    support[i] <- support_func(o, II[i,])
}
```

```{r}
x_try
length(support)
support <- support-max(support)
support <- pmax(support, -10)
support  <- matrix(support,length(r_try),length(x_try))
contour(r_try, x_try, support, xlab="r", ylab="x", levels = -c(1,(1:5)*2), lwd=c(1,3,1,1,1,1))
abline(v=1:9, lwd=0.3, col='gray')
abline(v = r_true)
abline(v = 1:9, col = 'gray', lty = 3, lwd = 0.3)
filled.contour(r_try, x_try, support, xlab = "r", ylab = "x")
```

```{r contourLO}
x_try
wanted <- 8:18
LO <- function(p){log(p/(1-p))}
contour(r_try, LO(x_try[wanted]), support[,wanted], xlab="r", ylab="Log-odds of x", levels = -c(1,(1:5)*2), lwd=c(1,3,1,1,1,1))
```

### 2025-10-02 inst/hankin-exponential_BT.tex
more work

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Robin K. S. Hankin} %% comma-separated
\Plaintitle{A Simplified Plackett-Luce Likelihood Function for Rank Orders}

@@

Here a new simplification of the Plackett--Luce likelihood functions
for order statistics is proposed.  We consider a field of $n$ entities
that are ordered in some way [the preferred example is competitors in
  a running race, ordered by time of crossing the finishing line].
One entity is of particular interest to us: the ``focal entity'', or
focal competitor.  We assign Plackett--Luce strength $a$ to the focal
competitor and $b$ to all others; we require $a+b=1$.  The associated
inference problem has a wide range of applications and I present some
analysis of datasets drawn from the fields of amateur athletics,
education, and Formula~1 motor racing.

}
\Keywords{keywords, comma-separated, non-capitalized, \proglang{R}}
\Plainkeywords{keywords, comma-separated, non-capitalized, R} %% without formatting
%% at least one keyword must be supplied

@@

>>>>

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Robin K. S. Hankin} %% comma-separated
\Plaintitle{A Parametrized Plackett-Luce Likelihood Function for Rank Orders}

@@

Here a new simplification of the Plackett--Luce likelihood functions
for order statistics is proposed.  We consider a field of $n$ entities
that are ordered in some way [the preferred example is competitors in
  a running race, ordered by time of crossing the finishing line].  We
assign Plackett--Luce strengths $\alpha x,\alpha x^2\ldots,\alpha x^n$
where $0\leq x\leqslant 1$ is a common ratio and $\alpha$ a
normalising constant.  One entity is of particular interest to us: the
``focal entity'', or focal competitor whose strength is $\alpha x^r$
for some unknown value of $r$ with $1\leqslant r\leqslant n$.  We
consider the following inference problem: given the order statistic
for the focal competitor [first, second, etc], and the total number of
competitors, what might we infer about $x$ and $r$?  Such problems
arise when non-focal competitors' performances are censored, unknown,
or irrelevant.  }

\Keywords{keywords, comma-separated, non-capitalized, \proglang{R}}
\Plainkeywords{keywords, comma-separated, non-capitalized, R} %% without formatting
%% at least one keyword must be supplied

@@

\subsection{Some simple cases}

We consider the smallest non-trivial cases $n=3$ and $n=4$.  For $n=3$
we have Plackett-Luce strengths $\alpha x,\alpha x^2,\alpha x^3$ [the
  extra factor of $x$ allows us to identify the $r$ in ``$\alpha x^r$''
  with the rank of the competitor, the best being rank 1].  Suppose
the focal competitor came first, the likelihood function would be

\begin{equation}
  \mathcal{L}(x,r)=\frac{x^{r-1}}{1+x+x^2}\qquad r=1,2,3
\end{equation}

If the focal competitor came second we would have two Plackett-Luce
probabilities to add, corresponding to $\alpha x^i$ or $\alpha x^j$
coming first where $i,j\neq r$.  For example, if $r=1$ we would have
two finishing orders that are consistent with the observation: $2\succ
1\succ 3$ and $3\succ 1\succ 2$.  These are disjoint so the likelihood
will be proportional to

$$
\frac{x^2}{x+x^2+x^3}\cdot\frac{x}{x+x^3} + 
\frac{x^3}{x+x^2+x^3}\cdot\frac{x}{x+x^2}
$$

After simplification:

\begin{equation}
  \mathcal{L}(x,r)=
  \begin{cases}
    \frac{x  }{x+x^2+x^3}\left(\frac{x^2}{x+x^3}   + \frac{x^3}{x+x^2}\right)\qquad r=1\\
    \frac{x^2}{x+x^2+x^3}\left(\frac{x  }{x^2+x^3} + \frac{x^3}{x+x^2}\right)\qquad r=2\\
    \frac{x^3}{x+x^2+x^3}\left(\frac{x  }{x^2+x^3} + \frac{x^2}{x+x^3}\right)\qquad r=3
\end{cases}
\end{equation}

The $n=4$ case is harder as we potentially have six consistent
orderings for each observation.  As an example, if the focal
competitor comes third we have $1
